Reproducible Research - Course Project 1
========================================


Introduction

It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the "quantified self" movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

The data for this assignment can be downloaded from the course web site:

- Dataset: Activity monitoring data [52K]

The variables included in this dataset are:

- steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)
- date: The date on which the measurement was taken in YYYY-MM-DD format
- interval: Identifier for the 5-minute interval in which measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

Throughout your report make sure you always include the code that you used to generate the output you present. When writing code chunks in the R markdown document, always use echo = TRUE so that someone else will be able to read the code. 

```{r global_options, include=TRUE}
#global settings
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='./',
                      echo=TRUE, warning=FALSE, message=FALSE)
```


##Loading and preprocessing the data


1. Load the data (i.e. read.csv())
2. Process/transform the data (if necessary) into a format suitable for your analysis

```{r loading_data}
activity <- read.csv("activity.csv")

```


##What is mean total number of steps taken per day? (For this part of the assignment, ignoring the missing values in the dataset.)

1. Calculate the total number of steps taken per day
2. Make a histogram of the total number of steps taken each day
3. Calculate and report the mean and median of the total number of steps taken per day

```{r steps_per_day}
steps_per_day <- aggregate(steps ~ date, data = activity, sum, na.rm = TRUE)

hist(steps_per_day$steps, main="Total number of steps per day", breaks = 25,
     col = "grey", xlab="Steps per day", ylab="Frequency")

#the mean
mean_steps_per_day <- mean(steps_per_day$steps, na.rm = TRUE)

#the median
median_steps_per_day<- median(steps_per_day$steps, na.rm = TRUE)


```
The mean of the total steps per day is `r mean_steps_per_day`. The median of total steps per day is `r median_steps_per_day`.


##What is the average daily activity pattern?

1. Time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
2. 5-minute interval, on average across all the days in the dataset, containing the maximum number of steps

```{r daily_pattern}
steps_per_interval <- aggregate(steps ~ interval, data=activity, mean)

# time series plot 
plot(steps_per_interval$interval, steps_per_interval$steps, type='l', 
     main="Average number of steps taken, averaged accros all days - daily activity pattern", xlab="5 min intervals", 
     ylab="Average number of steps")

# interval with max number of steps
steps_per_interval[which.max(steps_per_interval$steps), ]

```
The interval with the highest number of steps is `r steps_per_interval[which.max(steps_per_interval$steps), ]$interval`.


##Imputing missing values

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.
3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r impute}
# number of NA values
sum(is.na(activity))

#impute the mean of the interval when the step is NA
activity_new <- activity # new dataset called imp
for (i in steps_per_interval$interval) {
    activity_new[activity_new$interval == i & is.na(activity_new$steps), ]$steps <- 
        steps_per_interval$steps[steps_per_interval$interval == i]
}

steps_per_day_new <- aggregate(steps ~ date, data = activity_new, sum, na.rm = TRUE)

hist(steps_per_day_new$steps, breaks = 25, 
     main = "Total number of steps per day with imputed values",
     col = "grey", xlab = "Steps", ylab="Frequency")

# mean with imuted values
new_mean <- mean(steps_per_day_new$steps)

#median with imputed values
new_median <- median(steps_per_day_new$steps)

```
The new mean and median are `r new_mean` and `r new_median`. The mean and the median are the same after we imputed values.  The difference between in the median and mean between the original data and the data with imputed values in negligibly small. 



##Are there differences in activity patterns between weekdays and weekends?

1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.

```{r patterns}

activity_new['type_of_day'] <- weekdays(as.Date(activity_new$date))
activity_new$type_of_day[activity_new$type_of_day  %in% c('Saturday','Sunday') ] <- "weekend"
activity_new$type_of_day[activity_new$type_of_day != "weekend"] <- "weekday"

activity_new$type_of_day <- as.factor(activity_new$type_of_day)

steps_per_criteria <- aggregate(steps ~ interval + type_of_day, activity_new, mean)

library(lattice)
xyplot(steps ~ interval | type_of_day, data = steps_per_criteria, type = "l", lwd = 2,
       layout = c(1, 2), 
       xlab = "5-minute interval", 
       ylab = "Average number of steps per type of day",
       main = "Average number of steps")
```
We clearly see that during the weekends the test subject were more active throut the day.  During the week days the test subject we active primarily in the mornig and not the whole morning. 

